\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-12 paper/manuscript.tex
update

\title{\textbf{Introducing the Choice-Confidence (CHOCO) Model for
Bimodal Scales Data: Application to the Relationship between Reality
Beliefs about AI-Generated Faces and Beauty}}

@@

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, leftrule=.75mm, bottomrule=.15mm, opacityback=0, arc=.35mm, colback=white, rightrule=.15mm, left=2mm, colframe=quarto-callout-note-color-frame]

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data\footnote{More specifically, that the outcome
  is distributed according to a Normal distribution which parameters are
  expressed as a linear function of the predictors.} - as this
assumption often does not hold true for many types of psychological
outcomes. For instance, reaction times typically exhibit skewed
distributions, choices can be represented as binary variables, and count
data consists of strictly positive integers. Applying models that
presume normality and model the ``mean'' of the outcome variable can
lead to misinterpretation and potentially misleading conclusions when
applied indiscriminately. It is thus important that psychologists use
models that can best describe (or generate) the data they collect, to
fully exploit them and bring more nuance and accuracy to their
conclusions.

@@

Using the open-accss data from Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}), we included all heterosexual and
bisexual (as these two groups did not seem to differ based on
preliminary analyses and were thus grouped to maximize power) male and
female participants, for a final sample of 141 participants (Mean age =
28.4, SD = 9.0, range: {[}19, 66{]}; Sex: 47.5\% females). For each
participant, we included only stimuli of the opposite gender (i.e., all
89 female faces for men and 20 male faces for women).

@@

@@

@@

>>>>

\title{\textbf{Introducing the Choice-Confidence (CHOCO) Model for
Bimodal Data from Subjective Ratings: Application to the Effect of
Attractiveness on Reality Beliefs about AI-Generated Faces}}

@@

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, colback=white, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm]

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-blanca2018current}{Blanca et al., 2018};
\citeproc{ref-cumming2014new}{Cumming, 2014};
\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data\footnote{More specifically, that the outcome
  is distributed according to a Normal distribution which parameters are
  expressed as a linear function of the predictors.} - as this
assumption often does not hold true for many types of psychological
outcomes. For instance, reaction times typically exhibit skewed
distributions, choices can be represented as binary variables, and count
data consists of strictly positive integers. Applying models that
presume normality and model the ``mean'' of the outcome variable can
lead to misinterpretation and potentially misleading conclusions when
applied indiscriminately. It is thus important that psychologists use
models that can best describe (or generate) the data they collect, to
fully exploit them and bring more nuance and accuracy to their
conclusions.

@@

Using the open-access data from Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}), we included all heterosexual and
bisexual (as these two groups did not seem to differ based on
preliminary analyses and were thus grouped to maximize power) male and
female participants, for a final sample of 141 participants (Mean age =
28.4, SD = 9.0, range: {[}19, 66{]}; Sex: 47.5\% females). For each
participant, we included only stimuli of the opposite gender (i.e., all
89 female faces for men and 20 male faces for women).

@@

The findings of Study 2 replicate and extend those of Study 1,
supporting the idea that facial attractiveness significantly shapes
participants' beliefs about reality beliefs pertaining to human faces.
This effect was particularly strong and robust among male participants.
For females, the effect was weaker in Study 1 but more pronounced in
Study 2, which might partly reflect increased statistical power.

Another possibility, however, is that the second sample's experimental
manipulation introduced conflicting cues (e.g., faces labelled
``AI-generated'') that could have disrupted intuitive heuristics,
thereby increasing reliance on attractiveness as a diagnostic cue,
particularly in ambiguous trials.

@@

\bibitem[\citeproctext]{ref-blanca2018current}
Blanca, M. J., Alarcón, R., \& Bono, R. (2018). Current practices in
data analysis procedures in psychology: What has changed?
\emph{Frontiers in Psychology}, \emph{9}, 2558.

@@

\bibitem[\citeproctext]{ref-cumming2014new}
Cumming, G. (2014). The new statistics: Why and how. \emph{Psychological
Science}, \emph{25}(1), 7--29.
